---
title: "Statistical Report"
subtitle: "Determinants of Belgian Household Energy Consumption"

author: |
  GROUP 6:  
  Morodoluwa Akin-Fajiye  
  Samer Mheissen  
  Blagovesta Hristoskova  
  Jeffrey Nunoo  
  Edoardo Pennesi  
  Muthoni Wachira

date: "`r Sys.Date()`"

output:
  pdf_document:
    toc: true
    number_sections: true
bibliography: references.bib
csl: apa.csl
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{float}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

# Libraries
library(gridExtra)
library(dplyr)
library(knitr)
library(kableExtra)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(corrplot)



```

\newpage

# Background

The goal of “*Carbon Neutrality*” is gaining increasing momentum in the European political arena amid substantial indicators of an incoming climate change which might have deleterious effects at the societal, economic and, most importantly, environmental level. Household energy consumption plays a pivotal role in overall social sustainability since it accounts for a substantial share of the total energy demand in most developed countries. Within this sector, electricity consumption provides a useful proxy for household energy demand and can be put into relationship to structural characteristics of the buildings as well as the socio-economic profile of their occupants.

Dwelling-related attributes such as floor area, dwelling type, construction period, and Energy Performance Certificate (EPC) labels are consistently identified as important predictors of electricity demand [@mcloughlin2013; @santin2009]. Intuitively, larger dwellings typically require more energy for lighting, heating, cooling, and appliance use, while older buildings are constructed with materials which provides poorer insulation and therefore larger heat dissipation. Nevertheless, these variables interacts with socio-economic factors. Larger households topically host a higher number of occupants, therefore they can share the consumption for heating, refrigerators and air conditioning lowering the per-capita consumption. Similarly, high EPC label buildings are prerogative of high-income classes which might use a larger number of appliances and indulge into energy-consuming behaviours going to erode the benefits of a properly insulated house in terms of energy savings [@aydin2017]. 

Using household electricity consumption as a proxy for energy demand, this cross-sectional study examines its relationship
with building characteristics (e.g. floor area, EPC label, construction era) and socio-economic factors (number of permanent residents and annual net household income) collecting data from 2500 households in Belgium. Importantly, the study aims at characterising the interplay between EPC labels and the other factors involved in energy consumptions to provide data-driven tools to decision-makers. 


# Data management

## Data Cleaning

After loading the data, three observations (rows 66, 36, and 135) were removed from the dataset according to the instructions. Subsequently, the dataset was cleaned by excluding observations with missing values in the variables EPC label (`epc_label`) and annual household electricity consumption (`annual_kwh`) (none of the variables displayed missingness above 5\%). This step ensures that the analysis is conducted only on complete cases for the key variables of interest. Missing values in other variables were retained (see appendix) without imputation. To inspect the distribution of electricity consumption and identify potential extreme values, the observations were sorted in decreasing order of annual electricity consumption (`annual_kwh`). Five observations (row 267, 1330, 1570, 369, 12) were two orders of magnitude larger than the others (172 to 312 times larger) and, therefore, were removed as were attributed to imputation error. 

```{r, echo=FALSE, include=FALSE, warning=F, message=F}
data.students <- read.table("data_students.txt",header=TRUE,sep=";")
df_clean <- data.students[-c(66, 36, 135), ]
df_clean <- df_clean %>%
     filter(!is.na(epc_label) & !is.na(annual_kwh))
head(df_clean[order(-df_clean$annual_kwh), ], 10)
df_clean <- df_clean %>%
    filter(annual_kwh <= 18000)
```

## Data Coding

Variables were coded according to their type: continuous (e.g., `annual_kwh`, `sq_meters`, `income_euro`), count (`occupancy_count`), ordinal (`epc_label`, `building_era`), categorical (`roof_color`), or binary (`pet_ownership`). Table 1 summarizes their descriptions and coding schemes.

```{r echo=FALSE}
data_dict <- data.frame(
  Name = c("annual_kwh",
           "sq_meters",
           "epc_label",
           "building_era",
           "income_euro",
           "occupancy_count",
           "dist_to_brussels",
           "roof_color",
           "pet_ownership"),
  
  Description = c("yearly electricity use",
                  "indoor living area",
                  "energy performance certificate",
                  "construction period",
                  "net household income",
                  "permanent residents",
                  "distance to Brussels center",
                  "roof tile color",
                  "owner has a pet"),
  
  Type_Coding = c("Continuous (kWh/year)",
                  "Continuous (m²)",
                  "Ordinal: A, B, ..., F",
                  "Ordinal: dummy coded; ref = Pre-1970",
                  "Continuous (€/year)",
                  "Count: 1, 2, ..., n",
                  "Continuous (km)",
                  "Categorical: e.g., red, black, other",
                  "Binary: 0 = No, 1 = Yes")
)

kable(data_dict, caption = "Data Dictionary") %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(2, width = "6cm") %>%  
  column_spec(3, width = "6cm")
```


# Descriptive Statistics

## Categorical variables

The distribution of EPC labels is relatively balanced across categories. Labels A (22.2%) and C (21.3%) are the most common, followed by B (17.6%) and D (15.7%). Lower-efficiency categories E occur less frequently, 13.6% and F 9.6% respectively. Buildings constructed between 1970 and 2000 represent 34.4% of the sample, pre-1970 buildings account for 33.1%, and post-2000 buildings make up 32.6%. Occupancy count goes from 1 to 7 with all counts accounting approximately 14-15%. Red roofs are the most common (35.2%), followed closely by black roofs (33.3%) and other colors (31.5%). Almost one in four (39.6%) owners has a pet, while 60.4% does not. None of the categories appear too sparse, suggesting that all variables can be retained for modelling without concerns about low cell counts. Overall, the categorical variables show meaningful variation across categories. EPC labels are generally balanced across the different categories, indicating that extremely efficient or inefficient homes are not rare in this data. Building era and roof color are also evenly distributed while pet ownership is generally more common.

*Edo's comment: i think the sentence below can be replaced by the actual description as in the paragraph above and tables will go in the appendix*
These summary tables should help show if the EPC labels are balanced or skewed, if the building eras arw clustered within one period, if the roof colours or pet ownership categories have enough variation to be meaningful predictors. We can also sue them to investigate whether any category has very low counts, which can affect model stability.

### Summary tables
```{r echo=FALSE, warning=F, message=F}
# Create summary tables for categorical variables
epc_summary <- df_clean %>%
  count(epc_label) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

building_era_summary <- df_clean %>%
  count(building_era) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

roof_color_summary <- df_clean %>%
  count(roof_color) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

pet_summary <- df_clean %>%
  count(pet_ownership) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

occupancy_count <- df_clean %>%
  count(occupancy_count) %>%
  mutate(percent = round(n / sum(n) * 100, 1))


kable(epc_summary, caption = "EPC Label Distribution") %>% kable_styling()
kable(building_era_summary, caption = "Building Era Distribution") %>% kable_styling()
kable(roof_color_summary, caption = "Roof Color Distribution") %>% kable_styling()
kable(pet_summary, caption = "Pet Ownership Distribution") %>% kable_styling()
kable(occupancy_count, caption = "Occupancy Count Distribution") %>% kable_styling()

#crosscat <- df_clean %>%
  #count(epc_label, building_era, roof_color, pet_ownership) %>%
  #arrange(desc(n))

# crosscat

```


### Summary plots

```{r barplots_cat, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Bar plots for non-continuous variables", fig.pos="H"}
cat_vars <- df_clean %>%
  select(epc_label, building_era, roof_color, pet_ownership, occupancy_count) %>%
  mutate(across(everything(), as.character)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Category"
  ) %>%
  group_by(Variable, Category) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Variable) %>%
  mutate(Percent = Count / sum(Count) * 100)

ggplot(cat_vars, aes(x = Category, y = Percent)) +
  geom_col(fill = "gray") +
  geom_text(aes(label = paste0(round(Percent, 1), "%")),
            vjust = -0.5, size = 3) +
  facet_wrap(~ Variable, ncol = 2, scales = "free_x") +
  labs(
    x = NULL,
    y = "Percentage",
    title = "Distribution of Categorical Variables"
  ) +
  theme_minimal()
```

## Continuous variables


### Summary tables

```{r echo=FALSE, warning=F, message=F}
summary_table <- df_clean %>%
  summarise(
    across(
      c(annual_kwh, sq_meters, dist_to_brussels, income_euro),
      list(
        Mean = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Median = ~median(.x, na.rm = TRUE),
        IQR = ~IQR(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm = TRUE),
        Max = ~max(.x, na.rm = TRUE)
      )
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("Variable", "Statistic"),
    names_pattern = "(.+)_(Mean|SD|Median|IQR|Min|Max)$"
  ) %>%
  pivot_wider(names_from = Variable, values_from = value)


kable(summary_table,
      digits = 2,
      caption = "Summary Statistics for Continuous Variables")
```

### Summary plots

```{r echo=FALSE, warning=F, message=F}
# ---------------
# BOX PLOTS
# ---------------

par(mfrow = c(2,2))  # 1 row, # 2 row, 2 cols

boxplot(df_clean$annual_kwh,
        main = "Annual Energy Consumption",
        ylab = "annual_kwh")

boxplot(df_clean$sq_meters,
        main = "Floor Area",
        ylab = "sq_meters")

boxplot(df_clean$dist_to_brussels,
        main = "Distance to Brussels",  
        ylab = "dist_to_brussels")

boxplot(df_clean$income_euro,
        main = "Income",
        ylab = "income_euro")

par(mfrow = c(1,1))  # reset plotting layout



# ---------------
# CORR PLOTS
# ---------------
vars <- df_clean[, c("annual_kwh", "sq_meters", "dist_to_brussels", "income_euro")]
cor_matrix <- cor(vars, use = "complete.obs")
corrplot(cor_matrix,
         method = "circle",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45)




# ---------------
# QQ PLOTS
# ---------------
par(mfrow = c(2,2))  # 2 row, 2 cols

qqnorm(df_clean$annual_kwh,
       main = "QQ Plot: Annual Energy Consumption")
qqline(df_clean$annual_kwh)

qqnorm(df_clean$sq_meters,
       main = "QQ Plot: Floor Area")
qqline(df_clean$sq_meters)

qqnorm(df_clean$dist_to_brussels,
       main = "QQ Plot: Distance to Brussels")
qqline(df_clean$dist_to_brussels)

qqnorm(df_clean$income_euro,
       main = "QQ Plot: Income")
qqline(df_clean$income_euro)

par(mfrow = c(1,1))  # reset layout
```

# Inferential Statistics

## Regression model

## DAGs

## Results table

| Term | Estimate | Std. Error | t value | Pr(>|t|) | adjusted p |
|------|----------|------------|---------|----------| ---------  |

- VIF analysis
- Cross validation analysis


## Diagnostic plots

- Residuals vs Fitted
- Normal Q–Q Plot
- Scale–Location Plot (Spread–Location)
- Residuals vs Leverage
- Influence (Cook’s Distance)

## Contrast analysis

- for research question 2 and 4

# Conclusions

## Research question 1

## Research question 2

## Research question 3

## Research question 4
