---
title: "Statistical Report"
subtitle: "Determinants of Belgian Household Energy Consumption"

author: |
  GROUP 6:  
  Morodoluwa Akin-Fajiye  
  Samer Mheissen  
  Blagovesta Hristoskova  
  Jeffrey Nunoo  
  Edoardo Pennesi  
  Muthoni Wachira

date: "`r Sys.Date()`"

output:
  pdf_document:
    toc: true
    number_sections: true
bibliography: references.bib
csl: apa.csl
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{float}
  - \usepackage{placeins}
  - \usepackage{caption}
  - \captionsetup[table]{width=\textwidth}
  
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

# Libraries
library(gridExtra)
library(dplyr)
library(knitr)
library(kableExtra)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(corrplot)
library(purrr)
library(ggfortify)
library(broom)
library(multcomp)
library(rlang)
library(gtsummary)
library(dagitty)
library(ggdag)
library(ggplot2)
library(patchwork)
```

\newpage

# Background

The goal of “*Carbon Neutrality*” is gaining increasing momentum in the European political arena amid substantial indicators of an incoming climate change which might have deleterious effects at the societal, economic and, most importantly, environmental level. Household energy consumption plays a pivotal role in overall social sustainability since it accounts for a substantial share of the total energy demand in most developed countries. Within this sector, kilowatt-hour (kWh) consumption provides a useful measurement for overall household energy demand and can be put into relationship to structural characteristics of the buildings as well as the socio-economic profile of their occupants.

Dwelling related attributes such as floor area, dwelling type, construction period, and Energy Performance Certificate (EPC) labels are consistently identified as important predictors of electricity demand [@mcloughlin2013; @santin2009]. Intuitively, larger dwellings typically require more energy for lighting, heating, cooling, and appliance use, while older buildings are constructed with materials which provides poorer insulation and therefore larger heat dissipation. Nevertheless, these variables interacts with socio-economic factors. Larger households topically host a higher number of occupants, therefore they can share the consumption for heating, refrigerators and air conditioning and, consequently, lowering the *per capita* consumption. Similarly, one might speculate that high-class EPC label buildings are mostly prerogative of high-income classes which might use a larger number of appliances and indulge into energy-consuming behaviours going to erode the benefits of a properly insulated house in terms of absolute energy savings [@aydin2017]. This cross-sectional study collected data from 2500 households in Belgium and examined the relationship between households annual kWh consumption and several predictors using a linear regression model.


# Data management

## Data Cleaning

Three observations (rows 66, 36, and 135) were removed from the dataset according to the protocol The dataset was subsequently restricted to complete cases by excluding all observations (rows) with missing values, allowing estimation of the regression \(\beta\) coefficients via the least squares method. This approach relies on the assumption that the data are missing at random (MAR). None of the variables displayed missingness above 5% (see table 4). Five observations in the `annual_kwh` (rows 267, 1330, 1570, 369, 12) variable displayed values two orders of magnitude larger than the others (172 to 312 times larger) despite being comparable to other records across the other variables (indoor surface, EPC labe, etc...). Therefore, these were also removed as were considered implausible and likely attributed to imputation error (see section 6.2 for details). 

Categorical variables were encoded as factors, and reference levels were specified. Namely, the reference levels were defined as *"Pre-1970"* for `building_era`, *"A"* for `epc_label`, *"No"* for `pet_ownership` and *"Red"* for `roof_color`, against which all other categories were compared.

```{r, echo=FALSE, include=FALSE, warning=F, message=F}

# creating the dataset
data.students <- read.table("data_students.txt",header=TRUE,sep=";")
df_clean <- data.students[-c(66, 36, 135), ] # removing obs per protocol
df_clean <- df_clean %>%
     filter(if_all(everything(), ~ !is.na(.))) # missing vals removed
head(df_clean[order(-df_clean$annual_kwh), ], 10)
df_clean <- df_clean %>%
    filter(annual_kwh <= 18000) # outliers in kWh removed

# Coding the variables
df_clean$epc_label <- factor(df_clean$epc_label)
df_clean$building_era <- factor(df_clean$building_era)
df_clean$roof_color <- factor(df_clean$roof_color)
df_clean$pet_ownership <- factor(df_clean$pet_ownership)

# Setting the reference levels
df_clean$building_era <- relevel(df_clean$building_era, ref = "Pre-1970")
df_clean$epc_label <- relevel(df_clean$epc_label, ref = "A")
df_clean$roof_color <- relevel(df_clean$roof_color, ref = "Red")
df_clean$pet_ownership <- relevel(df_clean$pet_ownership, ref = "0")
```


# Descriptive Statistics

## Categorical variables

The distribution of EPC labels is relatively balanced across categories. Labels A (22.2%) and C (21.3%) are the most common, followed by B (17.6%) and D (15.7%). Lower-efficiency categories E occur less frequently, 13.6% and F 9.6% respectively. Buildings constructed between 1970 and 2000 represent 34.4% of the sample, pre-1970 33.1%, and post-2000 32.6%. Occupancy count ranges from 1 to 7 with all counts accounting for approximately 14-15%. A breakdown is provided in table 2. None of the categories appears too sparse, suggesting that all variables can be retained for modelling without concerns about low cell counts. From the exploratory analysis, three categorical variables seem to correlate with annual Kwh consumption: EPC label, building era and and occupancy count (figures 4-5). The other categorical variables do not display an apparent strong correlation with electricity consumption with similar means across each level. Moreover, EPC label and building era are correlated with each other since the most efficient labels are clustered in the two most recent building eras (figure 7). 


## Continuous variables

On average, households consume 5,874.76 kWh annually (considering the 5 outliers removed), with a relatively large standard deviation (2224.25 kWh). The median consumption (5432.07 kWh) is slightly lower than the mean, suggesting a right-skewed distribution, which is also supported by the QQ plot showing upward deviations above the diagonal line in the upper tail (figure 3). Floor area averages 158.25 m² with a standard deviation of 36.50 m², and a range from 60.26 m² to 341.30 m². Household income averages €48842.90, with a standard deviation of €22534.97 and a range from €8487.91 to €206367.91. As expected for this type of variable the QQ plot shows strong upper-tail deviation, suggesting left skewness driven by high-income outliers. A breakdown for each variable is provided in table 3. 

The correlation matrix reveals that annual electricity consumption has a moderate positive correlation with floor area (r = 0.38) and with income (r = 0.29), suggesting that larger apartments and wealthier households tend to consume more energy. Importantly, two predictors exhibit a strong positive correlation with each other, namely floor area and income (r = 0.72). Controlling for the income level, the EPC label distribution appears similar across income quantiles (figure 8) and comparable to the overall EPC distribution in the aggregate dataset. Indoor floor area does not associate with the size of the household since median floor area and interquartile ranges remain relatively stable when controlling for permanent residents, suggesting that larger households do not systematically live in larger dwellings in this dataset (figure 9). Distance to Brussels shows virtually no correlation with the other variables, suggesting that location relative to Brussels does not predict household energy use, dwelling size, or income in this dataset. More details are provided in sections 6.4 and 6.5 of the appendix.


# Inferential Statistics

## Regression model

The initial model contained all variables plus interaction terms between income and number of occupants, EPC label and floor area, EPC label and
number of occupants, and income and EPC label. Interactions were selected based on literature and protocol research questions. A quadratic term was considered for `occupancy_count` consistently with a possible decreasing marginal effect of additional occupants. We will refer to the initial model as *saturated*. 

Given that the primary objective of this study was to identify the set of regressors that best predict energy demand, explanatory variables retention was determined using a backward stepwise procedure based on the Bayesian Information Criterion (BIC) rather then exclusively on p-values. BIC favours more parsimonious models therefore lowering the risk of over-fitting. At each step, the variable whose removal produced the greatest reduction in BIC was eliminated, and the process continued until no further decrease was observed. 

The outcome variable `annual_kwh` was initially regressed using row data. Diagnostic plots (fig. 12) showed some violations of the standard assumptions. In particular, the residuals versus fitted plot indicates the presence of heteroskedasticity, with the variance of the residuals increasing for higher fitted values, and mild deviations from normality in the Q-Q plot. The outcome variable `annual_kwh` was then log-transformed resulting in an improvement of the diagnostic plots, particularly by yielding a more symmetric residual distribution and enhancing homoskedasticity. Log-transformed `annual_kwh` was used therafter for all model testing rpocedurs.

The final model retained floor area, EPC label, and the number of occupants as predictors of annual electricity consumption. The interaction terms and other covariates were removed. Nevertheless, `building_era` was retained in the model as it acts as a confounder being related to both `annual_kwh` and `epc_label`. We posit that building era affects energy consumption but also determines the energy calss since EPC label A and B were basically absent before year 2000 (see section 6.4 for details). 

Income was not retained in the final model and was not statistically significant (see Table 8). A plausible explanation is that the effect of income on energy consumption operates indirectly through dwelling size. As shown in Table 7, mean energy consumption across income quantiles becomes similar once stratifying by square meter quantiles. Moreover, income and dwelling size are strongly positively correlated. This suggests that dwelling size may act as a mediator in the relationship between income and energy consumption, capturing much of the variability that would otherwise be attributed to income.

Regression results for the saturated model, including p-values, are displayed in table 6. Based on statistical significance only, similar conclusions would have been achieved. Floor area, EPC label, household income, and occupancy count as predictors were the statistically significant ones. No evidence was found for non-linear effects of occupancy or for interaction effects, supporting a parsimonious model as the one selced using BIC criterion.

The results indicate that household size is a strong predictor of energy consumption, with each additional square meter associated with an approximate 0.4% increase in annual consumption. Assuming an average bed room size of 15 squared meters, each extra bedroom consumes on average 6% more energy. Compared to dwelling in energy class A, lower efficiency classes exhibit progressively higher electricity consumption, from about 7% for class B to nearly 48% for class F. 

Each additional occupant is responsible for around 16% increase in electricity consumption. No evidence of a decreasing marginal effect of the number of occupants was found at 5% significance level. Since there is no evidence that larger households tend to live in larger dwellings, a plausible explanation for the observed linear relationship between energy consumption and the number of occupants is that energy demand is primarily driven by individual behaviours, such as water use and appliance use, which are largely non-shareable. 

The final model achieved a substantially lower BIC (table 7) while maintaining nearly identical R² (89%), indicating that almost all the variance in annual consumption is explained by EPC label, house size and number of occupants. Overall, the diagnostic plots indicate that model assumptions are reasonably satisfied, with only minor deviations that are unlikely to substantially affect inference. the residuals vs leverage plot shows a few observations with higher leverage, but none appear to exert undue influence based on Cook’s distance, indicating the absence of strongly influential points.


```{r echo=FALSE, warning=F, message=F, include=FALSE}
# testing the initial saturated model with all vars and interactions
full_model <- lm(
  annual_kwh ~ sq_meters +
    epc_label +
    building_era +
    income_euro +
    occupancy_count +
    I(occupancy_count^2) +
    dist_to_brussels +
    roof_color +
    pet_ownership +
    income_euro:occupancy_count +
    income_euro:sq_meters+
    income_euro:epc_label+
    epc_label:sq_meters +
    epc_label:occupancy_count,
  data = df_clean
)


n <- nrow(df_clean)

final_model <- step(
  full_model,
  direction = "backward",
  k = log(n) # BIC criterion used
)

```

```{r echo=FALSE, warning=F, message=F, include=FALSE}
# visualise summary results (not displayed in the pdf)
summary(final_model, test=adjusted("BH"))
BIC(final_model)
```


```{r echo=FALSE, warning=F, message=F, include=FALSE}
# testing the initial saturated model with all vars and interactions but log_kWh
full_model.log <- lm(
  log(annual_kwh) ~ sq_meters +
    epc_label +
    building_era +
    income_euro +
    occupancy_count +
    I(occupancy_count^2) +
    dist_to_brussels +
    roof_color +
    pet_ownership +
    income_euro:occupancy_count +
    income_euro:sq_meters+
    income_euro:epc_label+
    epc_label:sq_meters +
    epc_label:occupancy_count,
  data = df_clean
)

n <- nrow(df_clean)

final_model.log <- step(
  full_model.log,
  direction = "backward",
  k = log(n)
)

# visualise summary results (not displayed in the pdf)
summary(final_model.log, test=adjusted("BH"))
BIC(final_model.log)
```




```{r echo=FALSE, warning=F, message=F, include=FALSE}
# Forcing building era into the model as a confounder

final_model.log_buil.era <- lm(
  log(annual_kwh) ~ sq_meters +
    epc_label +
    building_era+
    occupancy_count,
  data = df_clean
)

# visualise summary results (not displayed in the pdf)
summary(final_model.log_buil.era, test=adjusted("BH"))
BIC(final_model.log_buil.era)
```



```{r echo=FALSE, warning=F, message=F}
# CREATE A SUMMARY TABLE WITH REG RESULTS

# extract coefficients
coef_table <- summary(final_model.log_buil.era)$coefficients
ci <- confint(final_model.log_buil.era)

# Combine results
coef_table <- cbind(coef_table, ci)

# Convert to percentage effect on kWh
perc_effect <- (exp(coef_table[, "Estimate"]) - 1) * 100

# Transform CI
perc_ci_low  <- (exp(coef_table[, "2.5 %"]) - 1) * 100
perc_ci_high <- (exp(coef_table[, "97.5 %"]) - 1) * 100

# Keep selected columns (dropped beta CI and t statistics)
coef_table <- cbind(
  coef_table[, c("Estimate", "Std. Error", "Pr(>|t|)")],
  perc_effect, perc_ci_low, perc_ci_high
)

# Rename columns
colnames(coef_table) <- c(
  "Beta", "Std. Error", "p-value",
  "% Effect", "CI 2.5%", "CI 97.5%"
)

# Rename long row names
rownames(coef_table)[rownames(coef_table) == "building_era1970-2000"] <- "Built 1970–2000"
rownames(coef_table)[rownames(coef_table) == "building_eraPost-2000"] <- "Built post-2000"

kable(coef_table,
      digits = 3,
      caption = "Regression results for the final model (p-values are adjusted using the Benjamini-Hochberg procedure)")
```

```{r diag.plots, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Diagnostic plots for the selected regression model using log-transformed annual energy consuption.", fig.width=7, fig.height=7, fig.show="hold"}

# show diagnostic plots for the final model
par(mfrow = c(2,2))
plot(final_model.log_buil.era)
par(mfrow = c(1,1))

```

# Conclusions

The present analysis shows that, as already confirmed by literature, EPC label and number of occupants are the most relevant predictors of annual electricity consumption. Compared to dwelling in energy class A, lower efficiency classes exhibit progressively higher electricity consumption. Building in class F consumes almost 50% more than those in class A. Income does not appear to have a direct effect on energy consumption once dwelling characteristics are accounted for. A possible explanation is that income operates indirectly through larger living spaces (square_meters). There is no evidence that the effect of EPC label varies with household occupancy. Interaction terms between EPC label and occupancy were not retained in the final model and were not statistically significant.

From the policy-maker perspective, data suggest that interventions aimed at reducing residential energy demand should prioritise improving building energy performance. It remains to be established whether encouraging the use of common areas (for example co-working spaces in the building) will impact substantially on total energy demand. 


\newpage

# Appendix

## Summary tables for non-continuous variables

```{r echo=FALSE, warning=F, message=F}
# Create summary tables for categorical variables
cat_summary <- df_clean %>%
  select(where(is.factor)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable, value) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(variable) %>%
  mutate(p = round(100 * n / sum(n), 1))

knitr::kable(cat_summary, caption = "Categorical variables (n, %)")

```


## Summary tables for continuous varibales

```{r echo=FALSE, warning=F, message=F}
# Create summary tables for continuous variables
summary_table <- df_clean %>%
  summarise(
    across(
      c(annual_kwh, sq_meters, dist_to_brussels, income_euro),
      list(
        Mean = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Median = ~median(.x, na.rm = TRUE),
        IQR = ~IQR(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm = TRUE),
        Max = ~max(.x, na.rm = TRUE)
      )
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("Variable", "Statistic"),
    names_pattern = "(.+)_(Mean|SD|Median|IQR|Min|Max)$"
  ) %>%
  pivot_wider(names_from = Variable, values_from = value)

kable(summary_table,
      digits = 2,
      caption = "Summary Statistics for Continuous Variables")
```
\newpage

## Summary tables for missing values

```{r echo=FALSE, warning=F, message=F}

missing_table <- data.students %>%
  summarise(across(everything(),
    ~ sum(is.na(.))
  )) %>%
  pivot_longer(everything(),
               names_to = "Variable",
               values_to = "Missing_n") %>%
  mutate(
    Missing_pct = round(100 * Missing_n / nrow(df_clean), 2)
  ) %>%
  arrange(desc(Missing_pct))

knitr::kable(missing_table, caption = "Missing data summary")
```

## Summary plots for continuous varibales:exploratory analysis

Figures below refer to the dataset used for the regression model from which missing values and outliers in annual energy consumption were removed as explained in section 2 From the box plot we observe that several variables exhibit right-skewness and upper-tail outliers, particularly income and energy consumption. While annual energy consumption deviates significantly from normality, its log transformed analogue follows an approximate normal distribution dispute a mild deviation from the expected quantiles in the lower and upper tail. The QQ plot for floor area shows only mild departures from normality. The average distance to Brussels is 62.96 km. Its QQ plot indicates noticeable curvature, implying that the distribution deviates from normality, possibly reflecting clustering of households at certain distance ranges during the sampling procedure. Household income QQ plot shows strong upper-tail deviation, suggesting positive skewness driven by high-income outliers.   


```{r boxplots_cont, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Box plots for continuous variables", fig.width=10, fig.height=8}

# Box plots

layout(matrix(1:4, 2, 2))

boxplot(df_clean$annual_kwh,
        main = "Annual Energy Consumption",
        ylab = "annual_kwh")

boxplot(df_clean$sq_meters,
        main = "Floor Area",
        ylab = "sq_meters")

boxplot(df_clean$dist_to_brussels,
        main = "Distance to Brussels",  
        ylab = "dist_to_brussels")

boxplot(df_clean$income_euro,
        main = "Income",
        ylab = "income_euro")

```


```{r QQplots_cont, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="QQ plots for continuous variables", fig.width=12, fig.height=15}

# QQ Plots

layout(matrix(1:4, 2, 2))


qqnorm(df_clean$annual_kwh,
       main = "QQ Plot: Annual Energy Consumption")
qqline(df_clean$annual_kwh)

qqnorm(df_clean$sq_meters,
       main = "QQ Plot: Floor Area")
qqline(df_clean$sq_meters)

qqnorm(df_clean$dist_to_brussels,
       main = "QQ Plot: Distance to Brussels")
qqline(df_clean$dist_to_brussels)

qqnorm(df_clean$income_euro,
       main = "QQ Plot: Income")
qqline(df_clean$income_euro)

```



```{r QQplots_log.cont, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="QQ plots for log-transformed continuous variables", fig.width=12, fig.height=15, fig.show="hold"}

# QQ Plots for log_kWh

layout(matrix(1:4, 2, 2))

qqnorm(log(df_clean$annual_kwh),
       main = "QQ Plot: Log Annual Energy Consumption")
qqline(log(df_clean$annual_kwh))

qqnorm(log(df_clean$sq_meters),
       main = "QQ Plot: Log Floor Area")
qqline(log(df_clean$sq_meters))

qqnorm(log(df_clean$dist_to_brussels),
       main = "QQ Plot: Log Distance to Brussels")
qqline(log(df_clean$dist_to_brussels))

qqnorm(log(df_clean$income_euro),
       main = "QQ Plot: Log Income")
qqline(log(df_clean$income_euro))

```



\newpage

## Relationship between annual energy consumption and non-continuous variables: exploratory analysis

Among categorical and ordinal variables, EPC label and occupancy count are the most correlated to annual energy consumption. In addition, EPC label and building era are correlated with each other since the most efficient labels are clustered in the two most recent building eras and most recent buildings. Therefore, building era might play as a confounder in the model.


```{r boxplots eng.cons vs cat_vars, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Box plots for annual energy consumption vs categorical variables (part 1)", fig.pos="H", fig.width=8, fig.height=8, fig.show="hold"}

# Box plot eergy consumption vs cat variables

par(mfrow = c(2,1))

boxplot(annual_kwh ~ epc_label,
        data = df_clean,
        main = "Annual Energy Consumption by EPC Label",
        ylab = "annual_kwh",
        xlab = "epc_label")

boxplot(annual_kwh ~ occupancy_count,
        data = df_clean,
        main = "Annual Energy Consumption by Occupancy Counts",
        ylab = "annual_kwh",
        xlab = "Occupancy Counts")
```


```{r boxplots eng.cons vs cat_vars2, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Box plots for annual energy consumption vs categorical variables (part 2)", fig.pos="H", fig.width=8, fig.height=8, fig.show="hold"}

par(mfrow = c(2,1))

boxplot(annual_kwh ~ building_era,
        data = df_clean,
        main = "Annual Energy Consumption by Building Era",
        ylab = "annual_kwh",
        xlab = "building_era")


boxplot(annual_kwh ~ pet_ownership,
        data = df_clean,
        main = "Annual Energy Consumption by Pet Ownership",
        ylab = "annual_kwh",
        xlab = "Pet_Ownership")

```

```{r boxplots eng.cons vs cat_vars3, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Box plots for annual energy consumption vs categorical variables (part 3)", fig.pos="H", fig.width=8, fig.height=4, fig.show="hold"}

par(mfrow = c(1,1))

boxplot(annual_kwh ~ roof_color,
        data = df_clean,
        main = "Annual Energy Consumption by Roof Color",
        ylab = "annual_kwh",
        xlab = "Roof Color")
```

\clearpage

## Causal relationships among variables

Building era was treated as a confounder in the relationship between EPC label and electricity consumption since it is related to both, the main predictor and the outcome variable.

Income was treated as a mediator in the relationship between EPC electricity consumption and size of the households since people with higher income can afford larger dwellings which in turn consume more electricity. Indeed, the income and apartment size are strongly correlated (rho=0.72).

```{r Bar plots cat_vars vs cat_vars, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Bar plots predictors against each other (selected cases)", fig.pos="H", fig.width=7, fig.height=7}

# Bar plot of categorical vars against each-other

par(mfrow = c(2,1))

df_clean$building_era <- as.factor(df_clean$building_era)
df_clean$epc_label <- as.factor(df_clean$epc_label)
df_clean$occupancy_count <- as.factor(df_clean$occupancy_count)

cols2 <- gray.colors(length(levels(df_clean$epc_label)))

barplot(table(df_clean$epc_label, df_clean$building_era),
        beside = TRUE,
        main = "ECP label by Building Era",
        ylab = "count",
        xlab = "Building Era")

legend("topright",
       inset = c(-0.05, 0),   # move right
       legend = levels(df_clean$epc_label),
       fill = cols2,
       cex = 0.8,
       bty = "n",
       xpd = TRUE)   

cols3 <- gray.colors(length(levels(df_clean$building_era)))

barplot(table(df_clean$occupancy_count, df_clean$building_era),
        beside = TRUE,
        col = cols3,
        main = "Occupancy Count Distribution by Building Era",
        ylab = "Count",
        xlab = "Building Era")

legend("topright",
       inset = c(-0.05, 0),   # move right
       legend = levels(df_clean$occupancy_count),
       fill = cols3,
       cex = 0.8,
       bty = "n",
       xpd = TRUE)   


par(mfrow = c(1,1))

```


```{r Bar plots income_EPC, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Bar plots of EPC label class by Income percentile", fig.pos="H", fig.width=4, fig.height=4}

# Divide income by quantile
df_clean$income_q <- cut(
  df_clean$income,
  breaks = quantile(df_clean$income, probs = seq(0, 1, 0.2), na.rm = TRUE),
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4", "Q5")
)

# Table EPC labels distribution by income quintile
tab <- table(df_clean$income_q, df_clean$epc_label)
cols <- gray.colors(length(levels(df_clean$epc_label)))


bp <- barplot(t(tab),
              beside = TRUE,
              col = cols,
              main = "EPC Label Distribution by Income",
              ylab = "Count",
              xlab = "Income Quintile",
              names.arg = levels(df_clean$income_q))
        
legend("topright",
       inset = c(-0.1, 0),   # move right
       legend = levels(df_clean$epc_label),
       fill = cols,
       cex = 0.8,
       bty = "n",
       xpd = TRUE)    
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Divide sq_meters by quantile
df_clean$sq_meters_q <- cut(
  df_clean$sq_meters,
  breaks = quantile(df_clean$sq_meters, probs = seq(0, 1, 0.2), na.rm = TRUE),
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4", "Q5")
)

```



```{r Box plots sq_meters_oocupancy, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Box plots of dwelings square meters by occupancy count", fig.pos="H", fig.width=7, fig.height=4}

# Visualize sq_meters by occupancy_count
df_clean$occupancy_count <- as.factor(unlist(df_clean$occupancy_count))
boxplot(sq_meters ~ occupancy_count,
        data = df_clean,
        main = "Dwelling Size by Occupants",
        xlab = "Occupancy count",
        ylab = "Square meters")

```

\clearpage

```{r echo=FALSE, warning=F, message=F}
# Table mean Annual Energy Consumption by Size and Income Quintiles
tab_mean <- tapply(df_clean$annual_kwh,
                   list(Size = df_clean$sq_meters_q,
                        Income = df_clean$income_q),
                   mean,
                   na.rm = TRUE)
kable(tab_mean,
      caption = "Mean Annual Energy Consumption by Size and Income Quintiles",
      digits = 0,
      row.names = TRUE) %>%
  kableExtra::add_header_above(c(" " = 1, "Income Quintiles" = ncol(tab_mean)))

```

```{r corr.plots_cont, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Correlation plots for continuous variables", fig.width=6, fig.height=6}

# Correlation plots of each continuous variable with eachother

par(mfrow = c(1,1))  # reset plotting layout

vars <- df_clean[, c("annual_kwh", "sq_meters", "dist_to_brussels", "income_euro")]
cor_matrix <- cor(vars, use = "complete.obs")
corrplot(cor_matrix,
         method = "circle",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45)

```


```{r DAGs, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Directed acyclic graphs (DAGs)", fig.width=9, fig.height=5}
# Create DAG for the causal relationship among predictors and aoutcome

# DAG 1: Confounder
dag_confounder <- dagitty("
dag {
  building_era -> epc_label
  building_era -> annual_kwh
  epc_label -> annual_kwh
}
")

p1 <- ggdag(dag_confounder, text = TRUE, use_labels = "name") +
  theme_dag() +
  ggtitle("Building Era as a Confounder")

# DAG 2: Mediator
dag_mediator <- dagitty("
dag {
  income -> sq_meters
  sq_meters -> annual_kwh
}
")

p2 <- ggdag(dag_mediator, text = TRUE, use_labels = "name") +
  theme_dag() +
  ggtitle("Square_meters as a Mediator")

# Combine plots
p1 + p2
```



\clearpage

## Regression model

The initial regression model was the following:

$$
\begin{aligned}
\text{annual\_kwh}_i =\;&
\beta_0
+ \beta_1 \,\text{sq\_meters}_i
+ \beta_2 \,\text{income\_euro}_i
+ \beta_3 \,\text{occupancy\_count}_i
+ \beta_4 \,\text{dist\_to\_brussels}_i \\
&+ \boldsymbol{\beta}_5 \,\text{epc\_label}_i
+ \boldsymbol{\beta}_6 \,\text{building\_era}_i
+ \boldsymbol{\beta}_7 \,\text{roof\_color}_i
+ \beta_8 \,\text{pet\_ownership}_i \\
&+ \beta_9 (\text{income\_euro}_i \times \text{occupancy\_count}_i) \\
&+ \boldsymbol{\beta}_{10} (\text{epc\_label}_i \times \text{sq\_meters}_i) \\
&+ \boldsymbol{\beta}_{11} (\text{epc\_label}_i \times \text{occupancy\_count}_i) \\
&+ \boldsymbol{\beta}_{12} (\text{epc\_label}_i \times \text{income\_euro}_i) \\
&+ \boldsymbol{\beta}_{13} (\text{occupancy\_count}_i^2)
+ \varepsilon_i
\end{aligned}
$$

The final model retained the following predictors:

$$
\begin{aligned}
\log\text{annual\_kwh}_i =\;&
\beta_0
+ \beta_1 \,\text{sq\_meters}_i
+ \beta_2 \,\text{occupancy\_count}_i
+ \beta_3 \,\text{epc\_label}_i
+ \beta_4 \,\text{building\_era}_i
+ \varepsilon_i
\end{aligned}
$$


## Summary tables of regression results for the saturated model 


```{r echo=FALSE, warning=F, message=F}
# Display reg results for the saturated model

coef_table <- summary(full_model.log)$coefficients
colnames(coef_table) <- c("Estimate", "Std. Error", "t value", "p-value")

kable(coef_table,
      digits = 5,
      caption = "Regression results for the saturated model using log-transformed outcome (p-values are adjusted using the Benjamini-Hochberg procedure).",
      booktabs = TRUE,
      escape = TRUE)
```


```{r echo=FALSE, warning=F, message=F, include=FALSE}
# Compare BIC, AIC and R2 for each model tested

models <- list(
  initial_model = full_model,
  initial_model.log = full_model.log,
  final_model = final_model.log_buil.era
)

data.frame(
  Model = names(models),
  AIC = sapply(models, AIC),
  BIC = sapply(models, BIC),
  R2 = sapply(models, function(m) summary(m)$r.squared)
)

model_comp <- data.frame(
  Model = names(models),
  AIC = sapply(models, AIC),
  BIC = sapply(models, BIC),
  R2 = sapply(models, function(m) summary(m)$r.squared)
)
```

```{r echo=FALSE, warning=F, message=F}
kable(model_comp, digits = 3, caption = "Model comparison (initial vs. final) using AIC, BIC, and R-squared")
```



\newpage

## Diagnostic plots 

**Model regressing row annual energy consumption** 

```{r diag.plots for model row.ann.cons, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Diagnostic plots for the saturated regression model using annual energy consuption without log transformation.", fig.width=9, fig.height=8, fig.show="hold"}

par(mfrow = c(2,2))
plot(full_model)
```

\newpage

**Model regressing log-transformed annual energy consumption** 

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Diagnostic plots for the saturated regression model using log-transformed annual energy consuption.", fig.width=9, fig.height=8, fig.show="hold"}

par(mfrow = c(2,2))
plot(full_model.log)
par(mfrow = c(1,1))

```


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Diagnostic plots for the selected regression model using log-transformed annual energy consuption.", fig.width=9, fig.height=8, fig.show="hold"}

par(mfrow = c(2,2))
plot(final_model.log_buil.era)
par(mfrow = c(1,1))

```

\newpage

## R codes

```{r, eval=FALSE}
# creating the dataset
data.students <- read.table("data_students.txt",header=TRUE,sep=";")
df_clean <- data.students[-c(66, 36, 135), ] # removing obs per protocol
df_clean <- df_clean %>%
     filter(if_all(everything(), ~ !is.na(.))) # missing vals removed
head(df_clean[order(-df_clean$annual_kwh), ], 10)
df_clean <- df_clean %>%
    filter(annual_kwh <= 18000) # outliers in kWh removed

# Coding the variables
df_clean$epc_label <- factor(df_clean$epc_label)
df_clean$building_era <- factor(df_clean$building_era)
df_clean$roof_color <- factor(df_clean$roof_color)
df_clean$pet_ownership <- factor(df_clean$pet_ownership)

# Setting the reference levels
df_clean$building_era <- relevel(df_clean$building_era, ref = "Pre-1970")
df_clean$epc_label <- relevel(df_clean$epc_label, ref = "A")
df_clean$roof_color <- relevel(df_clean$roof_color, ref = "Red")
df_clean$pet_ownership <- relevel(df_clean$pet_ownership, ref = "0")

# testing the initial saturated model with all vars and interactions
full_model <- lm(
  annual_kwh ~ sq_meters +
    epc_label +
    building_era +
    income_euro +
    occupancy_count +
    I(occupancy_count^2) +
    dist_to_brussels +
    roof_color +
    pet_ownership +
    income_euro:occupancy_count +
    income_euro:sq_meters+
    income_euro:epc_label+
    epc_label:sq_meters +
    epc_label:occupancy_count,
  data = df_clean
)


n <- nrow(df_clean)

final_model <- step(
  full_model,
  direction = "backward",
  k = log(n) # BIC criterion used
)


# visualise summary results (not displayed in the pdf)
summary(final_model, test=adjusted("BH"))
BIC(final_model)

# testing the initial saturated model with all vars and interactions but log_kWh
full_model.log <- lm(
  log(annual_kwh) ~ sq_meters +
    epc_label +
    building_era +
    income_euro +
    occupancy_count +
    I(occupancy_count^2) +
    dist_to_brussels +
    roof_color +
    pet_ownership +
    income_euro:occupancy_count +
    income_euro:sq_meters+
    income_euro:epc_label+
    epc_label:sq_meters +
    epc_label:occupancy_count,
  data = df_clean
)

n <- nrow(df_clean)

final_model.log <- step(
  full_model.log,
  direction = "backward",
  k = log(n)
)

# visualise summary results (not displayed in the pdf)
summary(final_model.log, test=adjusted("BH"))
BIC(final_model.log)

# Forcing building era into the model as a confounder

final_model.log_buil.era <- lm(
  log(annual_kwh) ~ sq_meters +
    epc_label +
    building_era+
    occupancy_count,
  data = df_clean
)

# visualise summary results (not displayed in the pdf)
summary(final_model.log_buil.era, test=adjusted("BH"))
BIC(final_model.log_buil.era)

# CREATE A SUMMARY TABLE WITH REG RESULTS

# extract coefficients
coef_table <- summary(final_model.log_buil.era)$coefficients
ci <- confint(final_model.log_buil.era)

# Combine results
coef_table <- cbind(coef_table, ci)

# Convert to percentage effect on kWh
perc_effect <- (exp(coef_table[, "Estimate"]) - 1) * 100

# Transform CI
perc_ci_low  <- (exp(coef_table[, "2.5 %"]) - 1) * 100
perc_ci_high <- (exp(coef_table[, "97.5 %"]) - 1) * 100

# Keep selected columns (dropped beta CI and t statistics)
coef_table <- cbind(
  coef_table[, c("Estimate", "Std. Error", "Pr(>|t|)")],
  perc_effect, perc_ci_low, perc_ci_high
)

# Rename columns
colnames(coef_table) <- c(
  "Beta", "Std. Error", "p-value",
  "% Effect", "CI 2.5%", "CI 97.5%"
)

# Rename long row names
rownames(coef_table)[rownames(coef_table) == 
                       "building_era1970-2000"] <- "Built 1970–2000"
rownames(coef_table)[rownames(coef_table) 
                     == "building_eraPost-2000"] <- "Built post-2000"

kable(coef_table,
      digits = 3,
      caption = "Regression results for the final model (p-values are adjusted 
      using the Benjamini-Hochberg procedure)")

# show diagnostic plots for the final model
par(mfrow = c(2,2))
plot(final_model.log_buil.era)
par(mfrow = c(1,1))

# Create summary tables for categorical variables
cat_summary <- df_clean %>%
  select(where(is.factor)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable, value) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(variable) %>%
  mutate(p = round(100 * n / sum(n), 1))

knitr::kable(cat_summary, caption = "Categorical variables (n, %)")

# Create summary tables for continuous variables
summary_table <- df_clean %>%
  summarise(
    across(
      c(annual_kwh, sq_meters, dist_to_brussels, income_euro),
      list(
        Mean = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Median = ~median(.x, na.rm = TRUE),
        IQR = ~IQR(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm = TRUE),
        Max = ~max(.x, na.rm = TRUE)
      )
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("Variable", "Statistic"),
    names_pattern = "(.+)_(Mean|SD|Median|IQR|Min|Max)$"
  ) %>%
  pivot_wider(names_from = Variable, values_from = value)

kable(summary_table,
      digits = 2,
      caption = "Summary Statistics for Continuous Variables")

missing_table <- data.students %>%
  summarise(across(everything(),
    ~ sum(is.na(.))
  )) %>%
  pivot_longer(everything(),
               names_to = "Variable",
               values_to = "Missing_n") %>%
  mutate(
    Missing_pct = round(100 * Missing_n / nrow(df_clean), 2)
  ) %>%
  arrange(desc(Missing_pct))

knitr::kable(missing_table, caption = "Missing data summary")

# Box plots

layout(matrix(1:4, 2, 2))

boxplot(df_clean$annual_kwh,
        main = "Annual Energy Consumption",
        ylab = "annual_kwh")

boxplot(df_clean$sq_meters,
        main = "Floor Area",
        ylab = "sq_meters")

boxplot(df_clean$dist_to_brussels,
        main = "Distance to Brussels",  
        ylab = "dist_to_brussels")

boxplot(df_clean$income_euro,
        main = "Income",
        ylab = "income_euro")

# QQ Plots

layout(matrix(1:4, 2, 2))


qqnorm(df_clean$annual_kwh,
       main = "QQ Plot: Annual Energy Consumption")
qqline(df_clean$annual_kwh)

qqnorm(df_clean$sq_meters,
       main = "QQ Plot: Floor Area")
qqline(df_clean$sq_meters)

qqnorm(df_clean$dist_to_brussels,
       main = "QQ Plot: Distance to Brussels")
qqline(df_clean$dist_to_brussels)

qqnorm(df_clean$income_euro,
       main = "QQ Plot: Income")
qqline(df_clean$income_euro)

# QQ Plots for log_kWh

layout(matrix(1:4, 2, 2))

qqnorm(log(df_clean$annual_kwh),
       main = "QQ Plot: Log Annual Energy Consumption")
qqline(log(df_clean$annual_kwh))

qqnorm(log(df_clean$sq_meters),
       main = "QQ Plot: Log Floor Area")
qqline(log(df_clean$sq_meters))

qqnorm(log(df_clean$dist_to_brussels),
       main = "QQ Plot: Log Distance to Brussels")
qqline(log(df_clean$dist_to_brussels))

qqnorm(log(df_clean$income_euro),
       main = "QQ Plot: Log Income")
qqline(log(df_clean$income_euro))


# Box plot eergy consumption vs cat variables

par(mfrow = c(2,1))

boxplot(annual_kwh ~ epc_label,
        data = df_clean,
        main = "Annual Energy Consumption by EPC Label",
        ylab = "annual_kwh",
        xlab = "epc_label")

boxplot(annual_kwh ~ occupancy_count,
        data = df_clean,
        main = "Annual Energy Consumption by Occupancy Counts",
        ylab = "annual_kwh",
        xlab = "Occupancy Counts")

par(mfrow = c(2,1))

boxplot(annual_kwh ~ building_era,
        data = df_clean,
        main = "Annual Energy Consumption by Building Era",
        ylab = "annual_kwh",
        xlab = "building_era")


boxplot(annual_kwh ~ pet_ownership,
        data = df_clean,
        main = "Annual Energy Consumption by Pet Ownership",
        ylab = "annual_kwh",
        xlab = "Pet_Ownership")

par(mfrow = c(1,1))

boxplot(annual_kwh ~ roof_color,
        data = df_clean,
        main = "Annual Energy Consumption by Roof Color",
        ylab = "annual_kwh",
        xlab = "Roof Color")


# Bar plot of categorical vars against each-other

par(mfrow = c(2,1))

df_clean$building_era <- as.factor(df_clean$building_era)
df_clean$epc_label <- as.factor(df_clean$epc_label)
df_clean$occupancy_count <- as.factor(df_clean$occupancy_count)

cols2 <- gray.colors(length(levels(df_clean$epc_label)))

barplot(table(df_clean$epc_label, df_clean$building_era),
        beside = TRUE,
        main = "ECP label by Building Era",
        ylab = "count",
        xlab = "Building Era")

legend("topright",
       inset = c(-0.05, 0),   # move right
       legend = levels(df_clean$epc_label),
       fill = cols2,
       cex = 0.8,
       bty = "n",
       xpd = TRUE)   

cols3 <- gray.colors(length(levels(df_clean$building_era)))

barplot(table(df_clean$occupancy_count, df_clean$building_era),
        beside = TRUE,
        col = cols3,
        main = "Occupancy Count Distribution by Building Era",
        ylab = "Count",
        xlab = "Building Era")

legend("topright",
       inset = c(-0.05, 0),   # move right
       legend = levels(df_clean$occupancy_count),
       fill = cols3,
       cex = 0.8,
       bty = "n",
       xpd = TRUE)   


par(mfrow = c(1,1))


# Divide income by quantile
df_clean$income_q <- cut(
  df_clean$income,
  breaks = quantile(df_clean$income, probs = seq(0, 1, 0.2), na.rm = TRUE),
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4", "Q5")
)

# Table EPC labels distribution by income quintile
tab <- table(df_clean$income_q, df_clean$epc_label)
cols <- gray.colors(length(levels(df_clean$epc_label)))


bp <- barplot(t(tab),
              beside = TRUE,
              col = cols,
              main = "EPC Label Distribution by Income",
              ylab = "Count",
              xlab = "Income Quintile",
              names.arg = levels(df_clean$income_q))
        
legend("topright",
       inset = c(-0.1, 0),   # move right
       legend = levels(df_clean$epc_label),
       fill = cols,
       cex = 0.8,
       bty = "n",
       xpd = TRUE)    

# Divide sq_meters by quantile
df_clean$sq_meters_q <- cut(
  df_clean$sq_meters,
  breaks = quantile(df_clean$sq_meters, probs = seq(0, 1, 0.2), na.rm = TRUE),
  include.lowest = TRUE,
  labels = c("Q1", "Q2", "Q3", "Q4", "Q5")
)


# Visualize sq_meters by occupancy_count
df_clean$occupancy_count <- as.factor(unlist(df_clean$occupancy_count))
boxplot(sq_meters ~ occupancy_count,
        data = df_clean,
        main = "Dwelling Size by Occupants",
        xlab = "Occupancy count",
        ylab = "Square meters")

# Table mean Annual Energy Consumption by Size and Income Quintiles
tab_mean <- tapply(df_clean$annual_kwh,
                   list(Size = df_clean$sq_meters_q,
                        Income = df_clean$income_q),
                   mean,
                   na.rm = TRUE)
kable(tab_mean,
      caption = "Mean Annual Energy Consumption by Size and Income Quintiles",
      digits = 0,
      row.names = TRUE) %>%
  kableExtra::add_header_above(c(" " = 1, "Income Quintiles" = ncol(tab_mean)))



# Correlation plots of each continuous variable with eachother

par(mfrow = c(1,1))  # reset plotting layout

vars <- df_clean[, c("annual_kwh", "sq_meters", "dist_to_brussels", 
                     "income_euro")]
cor_matrix <- cor(vars, use = "complete.obs")
corrplot(cor_matrix,
         method = "circle",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45)


# Create DAG for the causal relationship among predictors and aoutcome

# DAG 1: Confounder
dag_confounder <- dagitty("
dag {
  building_era -> epc_label
  building_era -> annual_kwh
  epc_label -> annual_kwh
}
")

p1 <- ggdag(dag_confounder, text = TRUE, use_labels = "name") +
  theme_dag() +
  ggtitle("Building Era as a Confounder")

# DAG 2: Mediator
dag_mediator <- dagitty("
dag {
  income -> sq_meters
  sq_meters -> annual_kwh
}
")

p2 <- ggdag(dag_mediator, text = TRUE, use_labels = "name") +
  theme_dag() +
  ggtitle("Square_meters as a Mediator")

# Combine plots
p1 + p2

# Display reg results for the saturated model

coef_table <- summary(full_model.log)$coefficients
colnames(coef_table) <- c("Estimate", "Std. Error", "t value", "p-value")

kable(coef_table,
      digits = 5,
      caption = "Regression results for the saturated model using l
      og-transformed outcome (p-values are adjusted using the 
      Benjamini-Hochberg procedure).",
      booktabs = TRUE,
      escape = TRUE)


# Compare BIC, AIC and R2 for each model tested

models <- list(
  initial_model = full_model,
  initial_model.log = full_model.log,
  final_model = final_model.log_buil.era
)

data.frame(
  Model = names(models),
  AIC = sapply(models, AIC),
  BIC = sapply(models, BIC),
  R2 = sapply(models, function(m) summary(m)$r.squared)
)

model_comp <- data.frame(
  Model = names(models),
  AIC = sapply(models, AIC),
  BIC = sapply(models, BIC),
  R2 = sapply(models, function(m) summary(m)$r.squared)
)

kable(model_comp, digits = 3, caption = "Model comparison (initial vs. final) 
      using AIC, BIC, and R-squared")

# Diagnostic plots for each model
par(mfrow = c(2,2))
plot(full_model)
plot(full_model.log)
plot(final_model.log_buil.era)
par(mfrow = c(1,1))
```

\newpage



# References
