---
title: "Statistical Report"
subtitle: "Determinants of Belgian Household Energy Consumption"

author: |
  GROUP 6:  
  Morodoluwa Akin-Fajiye  
  Samer Mheissen  
  Blagovesta Hristoskova  
  Jeffrey Nunoo  
  Edoardo Pennesi  
  Muthoni Wachira

date: "`r Sys.Date()`"

output:
  pdf_document:
    toc: true
    number_sections: true
bibliography: references.bib
csl: apa.csl
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{float}
  - \usepackage{placeins}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

# Libraries
library(gridExtra)
library(dplyr)
library(knitr)
library(kableExtra)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(corrplot)
library(purrr)
library(ggfortify)
library(broom)
library(multcomp)
```

\newpage

# Background

The goal of “*Carbon Neutrality*” is gaining increasing momentum in the European political arena amid substantial indicators of an incoming climate change which might have deleterious effects at the societal, economic and, most importantly, environmental level. Household energy consumption plays a pivotal role in overall social sustainability since it accounts for a substantial share of the total energy demand in most developed countries. Within this sector, electricity consumption provides a useful proxy for overall household energy demand and can be put into relationship to structural characteristics of the buildings as well as the socio-economic profile of their occupants.

Dwelling related attributes such as floor area, dwelling type, construction period, and Energy Performance Certificate (EPC) labels are consistently identified as important predictors of electricity demand [@mcloughlin2013; @santin2009]. Intuitively, larger dwellings typically require more energy for lighting, heating, cooling, and appliance use, while older buildings are constructed with materials which provides poorer insulation and therefore larger heat dissipation. Nevertheless, these variables interacts with socio-economic factors. Larger households topically host a higher number of occupants, therefore they can share the consumption for heating, refrigerators and air conditioning lowering the *per capita* consumption. Similarly, high EPC label buildings are prerogative of high-income classes which might use a larger number of appliances and indulge into energy-consuming behaviours going to erode the benefits of a properly insulated house in terms of energy savings [@aydin2017]. This cross-sectional study collected data from 2500 households in Belgium and examines the relationship between households electricity consumption and several predictors.


# Data management

## Data Cleaning

Three observations (rows 66, 36, and 135) were removed from the dataset according to the instructions. Subsequently, the dataset was cleaned by excluding observations with missing values in the variables EPC label (`epc_label`) and annual household electricity consumption (`annual_kwh`). This step ensures that the analysis is conducted only on complete cases for the key variables of interest. Missing values in other variables were retained (see appendix) without imputation. None of the variables displayed missingness above 5%. Five observations in the`annual_kwh` (row 267, 1330, 1570, 369, 12) variable displayed values two orders of magnitude larger than the others (172 to 312 times larger) and, therefore, were removed as were attributed to imputation error (see section 6.2 for details). 

```{r, echo=FALSE, include=FALSE, warning=F, message=F}
data.students <- read.table("data_students.txt",header=TRUE,sep=";")
df_clean <- data.students[-c(66, 36, 135), ]
df_clean <- df_clean %>%
     filter(!is.na(epc_label) & !is.na(annual_kwh))
head(df_clean[order(-df_clean$annual_kwh), ], 10)
df_clean <- df_clean %>%
    filter(annual_kwh <= 18000)
```

## Data Coding

Variables were coded according to their type: continuous, count, ordinal, categorical, or binary. Table 1 summarizes their descriptions and coding schemes.

\newpage

```{r echo=FALSE}
data_dict <- data.frame(
  Name = c("annual_kwh",
           "sq_meters",
           "epc_label",
           "building_era",
           "income_euro",
           "occupancy_count",
           "dist_to_brussels",
           "roof_color",
           "pet_ownership"),
  
  Description = c("yearly electricity use",
                  "indoor living area",
                  "energy performance certificate",
                  "construction period",
                  "net household income",
                  "permanent residents",
                  "distance to Brussels center",
                  "roof tile color",
                  "owner has a pet"),
  
  Type_Coding = c("Continuous (kWh/year)",
                  "Continuous (m²)",
                  "Ordinal: A, B, ..., F",
                  "Ordinal: ref = Pre-1970, 1970-2000, post-2000",
                  "Continuous (€/year)",
                  "Count: 1, 2, ..., n",
                  "Continuous (km)",
                  "Categorical: red, black, other",
                  "Binary: 0 = No, 1 = Yes")
)

kable(data_dict, caption = "Data Dictionary") %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(2, width = "6cm") %>%  
  column_spec(3, width = "6cm")
```


# Descriptive Statistics

## Categorical variables

The distribution of EPC labels is relatively balanced across categories. Labels A (22.2%) and C (21.3%) are the most common, followed by B (17.6%) and D (15.7%). Lower-efficiency categories E occur less frequently, 13.6% and F 9.6% respectively. Buildings constructed between 1970 and 2000 represent 34.4% of the sample, pre-1970 33.1%, and post-2000 32.6%. Occupancy count ranges from 1 to 7 with all counts accounting approximately 14-15%. A breakdown is provided in the appendix. None of the categories appear too sparse, suggesting that all variables can be retained for modelling without concerns about low cell counts. From the exploratory analysis, two categorical variables seem to correlate strongly with electricity consumption: EPC label and occupancy count. The other categorical variables do not display an apparent correlation with electricity consumption. Moreover, EPC label and building era are correlated with each other since the most efficient labels are clustered in the two most recent building eras. (see section 6.4). 


## Continuous variables

On average, households consume 5,874.76 kWh of electricity annually (outliers removed), with a relatively large standard deviation (2,224.25 kWh). The median consumption (5,432.07 kWh) is slightly lower than the mean, suggesting a right-skewed distribution, which is also supported by the QQ plot showing upward deviations above the diagonal line in the upper tail (see appendix). Floor area averages 158.25 m² with a standard deviation of 36.50 m², and a range from 60.26 m² to 341.30 m². Household income averages €48,842.90, with a standard deviation of €22,534.97 and a range from €8,487.91 to €206,367.91. As expected for this type of variable the QQ plot shows strong upper-tail deviation, suggesting positive skewness driven by high-income outliers. A breakdown for each variable is provided in section 6.2. 

The correlation matrix reveals that annual electricity consumption has a moderate positive correlation with floor area (r = 0.38) and with income (r = 0.29), suggesting that larger and wealthier households tend to consume more energy. Importantly, two predictors exhibit a strong positive correlation with each other, namely floor area and income (r = 0.72). Distance to Brussels shows virtually no correlation with the other variables, suggesting that location relative to Brussels does not strongly influence household energy use, dwelling size, or income in this dataset.




# Inferential Statistics

## Regression model

The initial model contained all variables plus interaction terms between income and number of occupants, EPC label and floor area, EPC label and
number of occupants, and income and EPC label. Interactions were selected based on literature and exploratory analysis. A quadratic term was considered for `occupancy_count` consistently with a possible decreasing marginal effect. Predictor retention was determined using a backward stepwise procedure based on the Bayesian Information Criterion (BIC), which favours more parsimonious models. At each step, the variable whose removal produced the greatest reduction in BIC was eliminated, and the process continued until no further decrease was observed. The outcome variable `annual_kwh` was initially regressed using row data. Diagnostic plots (see appendix) showed some violations of the standard assumptions. In particular, the residuals vs fitted plot indicates the presence of heteroskedasticity, with the variance of the residuals increasing for higher fitted values, and mild deviations from normality in the Q-Q plot. The outcome variable `annual_kwh` was then log transformed resulting an an improvement of the diagnostic plots, particularly by yielding a more symmetric residual distribution and enhancing homoskedasticity (see section 6.5).

The final model retained floor area, EPC label, and the number of occupants as predictors of annual electricity consumption. The interaction terms and other covariates were removed. Nevertheless, `building_era` was retained in the model as it acts as a confounded being related to both `annual_kwh` and `epc_label` (see section 6.4 for details). Regression results for the initial full model are reported in the apendix for comparison. 


```{r echo=FALSE, warning=F, message=F, include=FALSE}

df_clean$epc_label <- factor(df_clean$epc_label)
df_clean$building_era <- factor(df_clean$building_era)
df_clean$roof_color <- factor(df_clean$roof_color)
df_clean$pet_ownership <- factor(df_clean$pet_ownership)

df_clean$building_era <- relevel(df_clean$building_era, ref = "Pre-1970")
df_clean$epc_label <- relevel(df_clean$epc_label, ref = "A")
df_clean$roof_color <- relevel(df_clean$roof_color, ref = "Red")


full_model <- lm(
  annual_kwh ~ sq_meters +
    epc_label +
    building_era +
    income_euro +
    occupancy_count +
    I(occupancy_count^2) +
    dist_to_brussels +
    roof_color +
    pet_ownership +
    income_euro:occupancy_count +
    epc_label:sq_meters +
    epc_label:occupancy_count +
    income_euro:epc_label,
  data = df_clean
)


n <- nrow(df_clean)

final_model <- step(
  full_model,
  direction = "backward",
  k = log(n)
)

```

```{r echo=FALSE, warning=F, message=F, include=FALSE}

summary(final_model)

BIC(final_model)
```


```{r echo=FALSE, warning=F, message=F, include=FALSE}

full_model.log <- lm(
  log(annual_kwh) ~ sq_meters +
    epc_label +
    building_era +
    income_euro +
    occupancy_count +
    I(occupancy_count^2) +
    dist_to_brussels +
    roof_color +
    pet_ownership +
    income_euro:occupancy_count +
    epc_label:sq_meters +
    epc_label:occupancy_count +
    income_euro:epc_label,
  data = df_clean
)

n <- nrow(df_clean)

final_model.log <- step(
  full_model.log,
  direction = "backward",
  k = log(n)
)

summary(final_model.log)
BIC(final_model.log)
```




```{r echo=FALSE, warning=F, message=F, include=FALSE}
# ------------------------------------------------
# FORCING BUILDING ERA IN THE MODEL AS CONFOUNDER
# ------------------------------------------------

final_model.log_buil.era <- lm(
  log(annual_kwh) ~ sq_meters +
    epc_label +
    building_era+
    occupancy_count,
  data = df_clean
)

summary(final_model.log_buil.era)
BIC(final_model.log_buil.era)
```

The results indicate that household size is a strong predictor of electricity consumption, with each additional square meter associated with an approximate 0.38% increase in annual consumption. Compared to dwelling in energy class A, lower efficiency classes exhibit progressively higher electricity consumption, from about 7% for class B to nearly 37% for class F. Each additional occupant is responsible for around 15-16% increase in electricity consumption. No evidence a decreasing marginal effect of the number of occupants was found at 5% significance level. The final model achieved a substantially lower BIC while maintaining nearly identical R² (89%), indicating that almost all the variance in annual consumption is explained by EPC label, house size and number of occupants.

```{r echo=FALSE, warning=F, message=F}

coef_table <- summary(final_model.log_buil.era)$coefficients
ci <- confint(final_model.log_buil.era)

# Combine results
coef_table <- cbind(coef_table, ci)

# Rename columns
colnames(coef_table) <- c("Estimate", "Std. Error", "t value", "p-value", "CI 2.5%", "CI 97.5%")

kable(coef_table,
      digits = 5,
      caption = "Regression results for the final model (with 95% CI)")
```

## Contrast analysis

Comparison using Dunnett contrast to compare each label B-F to A, with Benjamini-Hochberg correction for multiple testing, shows that annual energy use is significantly higher in the lower EPC ratings, compared to the reference category. Homes with a 'B' rating use approximately 9 % more energy than homes with the reference label, while 'C', 'D', 'E' and 'F' use about 17%, 27%, 37% and 48% respectively when converting to percent differences. This demonstrates a clear and consistent relationship between lower EPC ratings and energy use.

```{r echo=FALSE, warning=F, message=F, include=FALSE}

m.mcp<-glht(final_model.log_buil.era,linfct=mcp(epc_label="Dunnett"))
summary(m.mcp,test=adjusted("BH"))
```


```{r echo=FALSE, warning=F, message=F}

contrast.results <- summary(m.mcp)
coef_table <- data.frame(
  Estimate = contrast.results$test$coefficients,
  "Std. Error" = contrast.results$test$sigma,
  "t value" = contrast.results$test$tstat,
  "p-value" = contrast.results$test$pvalues
)

ci <- confint(m.mcp)$confint

coef_table <- cbind(coef_table, ci)
coef_table <- coef_table[, -c(5, 8,9,10)]

colnames(coef_table) <- c("Estimate", "Std. Error", "t value", "p-value",
                         "CI 2.5%", "CI 97.5%")

kable(coef_table,
      digits = 4,
      caption = "Contrast results for the EPC Labels")
```

A model with an interaction term between occupancy count and a dummy variable for high-quality EPC label A-B was built to investigate whether the effect of EPC label is dependent on the number of occupants. The analysis shows that the interaction term is not statistically significant with a p-value of 0.77. This signifies that there the effect of a proper insulation on electricity consumption does not change with the number of occupants.

```{r echo=FALSE, warning=F, message=F, include=FALSE}
df_clean$epc_high <- ifelse(df_clean$epc_label %in% c('A','B'),1,0)

model_q4<-lm(log(annual_kwh)~occupancy_count*epc_high+sq_meters+building_era,df_clean)

q4.results <- summary(model_q4)
```

```{r echo=FALSE, warning=F, message=F}

q4.results <- summary(model_q4)

coef_table <- as.data.frame(q4.results$coefficients)

colnames(coef_table) <- c("Estimate", "Std. Error", "t value", "p-value")

kable(coef_table,
      digits = 4,
      caption = "Results Interaction between EPC Label and Occupancy Count")
```


# Conclusions

The present analysis shows that, as already confirmed by literature, EPC label and number of occupants are the most relevant predictors of annual electricity consumption. Compared to dwelling in energy class A, lower efficiency classes exhibit progressively higher electricity consumption, from about 7% for class B to nearly 37% for class F. Each additional occupant
is responsible for a linear increase in electricity consumption of about 16%. Among the possible shortcomings of this study is that it considers only electricity consumption while in the overall energy consumption also gas represent a relevant quota. Actually, most heating systems are still powered by gas (methane) and heating is probably the most "shared" consumption in a household. This might explain why the expected reduced marginal contribution of additional occupants in the house was not apparent in this analysis. We also found that the difference among EPC labels is not evenly spaced. The gains in energy saving are much more sublets between EPC label A and B than among C to E and again only small between E and F. Income seems to be relevant as a mediator. Richer families tend to buy larger apartments and better insulated. The latter factors influence the electricity consumption.   


\newpage

# Appendix

## Summary tables for non-continuous varibales

Tables below refer to the dataset with only rows 66, 36 135 removed.

```{r echo=FALSE, warning=F, message=F}
# Create summary tables for categorical variables
epc_summary <- data.students[-c(66, 36, 135), ] %>%
  count(epc_label) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

building_era_summary <- data.students[-c(66, 36, 135), ] %>%
  count(building_era) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

roof_color_summary <- data.students[-c(66, 36, 135), ] %>%
  count(roof_color) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

pet_summary <- data.students[-c(66, 36, 135), ] %>%
  count(pet_ownership) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

occupancy_count <- data.students[-c(66, 36, 135), ] %>%
  count(occupancy_count) %>%
  mutate(percent = round(n / sum(n) * 100, 1))


kable(epc_summary, caption = "EPC Label Distribution") %>% kable_styling()
kable(building_era_summary, caption = "Building Era Distribution") %>% kable_styling()
kable(roof_color_summary, caption = "Roof Color Distribution") %>% kable_styling()

```

\newpage

```{r echo=FALSE, warning=F, message=F}

kable(pet_summary, caption = "Pet Ownership Distribution") %>% kable_styling()
kable(occupancy_count, caption = "Occupancy Count Distribution") %>% kable_styling()
```

\newpage

## Summary tables for continuous varibales

Tables below refer to the dataset with only rows 66, 36 135 removed and the dataset after NA and outliers removed.

```{r echo=FALSE, warning=F, message=F}
summary_table <- data.students[-c(66, 36, 135), ] %>%
  summarise(
    across(
      c(annual_kwh, sq_meters, dist_to_brussels, income_euro),
      list(
        Mean = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Median = ~median(.x, na.rm = TRUE),
        IQR = ~IQR(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm = TRUE),
        Max = ~max(.x, na.rm = TRUE)
      )
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("Variable", "Statistic"),
    names_pattern = "(.+)_(Mean|SD|Median|IQR|Min|Max)$"
  ) %>%
  pivot_wider(names_from = Variable, values_from = value)

kable(summary_table,
      digits = 2,
      caption = "Summary Statistics for Continuous Variables (row data)")
```

```{r echo=FALSE, warning=F, message=F}
summary_table <- data.students[-c(66, 36, 135), ] %>%
  summarise(
    across(
      c(annual_kwh, sq_meters, dist_to_brussels, income_euro),
      list(
        Mean = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Median = ~median(.x, na.rm = TRUE),
        IQR = ~IQR(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm = TRUE),
        Max = ~max(.x, na.rm = TRUE)
      )
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("Variable", "Statistic"),
    names_pattern = "(.+)_(Mean|SD|Median|IQR|Min|Max)$"
  ) %>%
  pivot_wider(names_from = Variable, values_from = value)

kable(summary_table,
      digits = 2,
      caption = "Summary Statistics for Continuous Variables (cleaned dataset)")
```


\newpage

## Summary plots for continuous varibales

Figures below refer to the dataset used for the regression model from which missing values and outliers in annual energy consumption were removed as explained in section two. From the box plot we observe that several variables exhibit right-skewness and upper-tail outliers, particularly income and energy consumption. While annual energy consumption deviates significantly from normality, its log transformed analogue follows an approximate normal distribution dispute a mild deviation from the expected quantiles in the lower and upper tail. The QQ plot for floor area shows only mild departures from normality. The average distance to Brussels is 62.96 km. Its QQ plot indicates noticeable curvature, implying that the distribution deviates from normality, possibly reflecting clustering of households at certain distance ranges during the sampling procedure. Household income QQ plot shows strong upper-tail deviation, suggesting positive skewness driven by high-income outliers.   


```{r boxplots_cont, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Box plots for continuous variables", fig.width=10, fig.height=8}

# ---------------
# BOX PLOTS
# ---------------

layout(matrix(1:4, 2, 2))

boxplot(df_clean$annual_kwh,
        main = "Annual Energy Consumption",
        ylab = "annual_kwh")

boxplot(df_clean$sq_meters,
        main = "Floor Area",
        ylab = "sq_meters")

boxplot(df_clean$dist_to_brussels,
        main = "Distance to Brussels",  
        ylab = "dist_to_brussels")

boxplot(df_clean$income_euro,
        main = "Income",
        ylab = "income_euro")

```


```{r QQplots_cont, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="QQ plots for continuous variables", fig.width=12, fig.height=15}

# ---------------
# QQ PLOTS
# ---------------

layout(matrix(1:4, 2, 2))


qqnorm(df_clean$annual_kwh,
       main = "QQ Plot: Annual Energy Consumption")
qqline(df_clean$annual_kwh)

qqnorm(df_clean$sq_meters,
       main = "QQ Plot: Floor Area")
qqline(df_clean$sq_meters)

qqnorm(df_clean$dist_to_brussels,
       main = "QQ Plot: Distance to Brussels")
qqline(df_clean$dist_to_brussels)

qqnorm(df_clean$income_euro,
       main = "QQ Plot: Income")
qqline(df_clean$income_euro)

```



```{r QQplots_log.cont, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="QQ plots for log-transformed continuous variables", fig.width=12, fig.height=15, fig.show="hold"}

# ---------------
# QQ PLOTS LOG 
# ---------------

layout(matrix(1:4, 2, 2))

qqnorm(log(df_clean$annual_kwh),
       main = "QQ Plot: Log Annual Energy Consumption")
qqline(log(df_clean$annual_kwh))

qqnorm(log(df_clean$sq_meters),
       main = "QQ Plot: Log Floor Area")
qqline(log(df_clean$sq_meters))

qqnorm(log(df_clean$dist_to_brussels),
       main = "QQ Plot: Log Distance to Brussels")
qqline(log(df_clean$dist_to_brussels))

qqnorm(log(df_clean$income_euro),
       main = "QQ Plot: Log Income")
qqline(log(df_clean$income_euro))

```



\newpage

## Relationship between Annual Energy Consumption and non-continuous variables.

Among categorical and ordinal variables, EPC label and occupancy count are the most correlated to annual energy consumption. In addition, EPC label and building era are correlated with each other since the most efficient labels are clustered in the two most recent building eras and most recent buildings. Therefore, building era might play as a confounder in the model.


```{r boxplots eng.cons vs cat_vars, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Box plots for annual energy consumption vs categorical variables (part 1)", fig.pos="H", fig.width=8, fig.height=8, fig.show="hold"}

# ---------------
# BOX PLOTS
# ---------------

par(mfrow = c(2,1))

boxplot(annual_kwh ~ epc_label,
        data = df_clean,
        main = "Annual Energy Consumption by EPC Label",
        ylab = "annual_kwh",
        xlab = "epc_label")

boxplot(annual_kwh ~ occupancy_count,
        data = df_clean,
        main = "Annual Energy Consumption by Occupancy Counts",
        ylab = "annual_kwh",
        xlab = "Occupancy Counts")
```


```{r boxplots eng.cons vs cat_vars2, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Box plots for annual energy consumption vs categorical variables (part 2)", fig.pos="H", fig.width=8, fig.height=8, fig.show="hold"}

par(mfrow = c(2,1))

boxplot(annual_kwh ~ building_era,
        data = df_clean,
        main = "Annual Energy Consumption by Building Era",
        ylab = "annual_kwh",
        xlab = "building_era")


boxplot(annual_kwh ~ pet_ownership,
        data = df_clean,
        main = "Annual Energy Consumption by Pet Ownership",
        ylab = "annual_kwh",
        xlab = "Pet_Ownership")

```

```{r boxplots eng.cons vs cat_vars3, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Box plots for annual energy consumption vs categorical variables (part 3)", fig.pos="H", fig.width=8, fig.height=4, fig.show="hold"}

par(mfrow = c(1,1))

boxplot(annual_kwh ~ roof_color,
        data = df_clean,
        main = "Annual Energy Consumption by Roof Color",
        ylab = "annual_kwh",
        xlab = "Roof Color")
```

## Causal Relationships Among Variables

Building era was treated as a confounder in the relationship between EPC label and electricity consumption since it is related to both: the main predictor and the outcome variable.

Income was treated as a mediator in the relationship between EPC electricity consumption and size of the households since people with higher income can afford larger dwellings which in turn consume more electricity. Indeed, the income and apartment size are strongly correlated (rho=0.72).

```{r Bar plots cat_vars vs cat_vars, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Bar plots predictors against each other (selected cases)", fig.pos="H", fig.width=7, fig.height=7}

par(mfrow = c(2,1))

df_clean$building_era <- as.factor(df_clean$building_era)
df_clean$epc_label <- as.factor(df_clean$epc_label)
df_clean$occupancy_count <- as.factor(df_clean$occupancy_count)
cols2 <- gray.colors(length(levels(df_clean$occupancy_count)))

barplot(table(df_clean$epc_label, df_clean$building_era),
        beside = TRUE,
        legend = TRUE,
        main = "ECP label by Building Era",
        ylab = "epc_label",
        xlab = "count")

barplot(table(df_clean$occupancy_count, df_clean$building_era),
        beside = TRUE,
        col = cols2,
        main = "Occupancy Count Distribution by Building Era",
        ylab = "Count",
        xlab = "Building Era")

legend("topright",
       legend = levels(df_clean$occupancy_count),
       fill = cols2,
       horiz = TRUE,
       xpd = TRUE,
       cex = 0.8,
       bty = "n")


par(mfrow = c(1,1))

```


```{r Bar plots income_EPC, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Bar plots of EPC label class by Income percentile", fig.pos="H", fig.width=7, fig.height=7}

df_clean$income_q <- cut(
  df_clean$income,
  breaks = quantile(df_clean$income, probs = seq(0, 1, 0.2), na.rm = TRUE),
  include.lowest = TRUE,
  labels = c("Q1 (lowest)", "Q2", "Q3", "Q4", "Q5 (highest)")
)

cols <- gray.colors(length(levels(df_clean$epc_label)))

barplot(table(df_clean$income_q, df_clean$epc_label),
        beside = TRUE,
        col = cols,
        main = "EPC Label Distribution by Income Quintile",
        ylab = "Count",
        xlab = "Income Quintile")
        
legend("topright",
       legend = levels(df_clean$epc_label),
       fill = cols,
       cex = 0.8,
       bty = "n")
```


```{r corr.plots_cont, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Correlation plots for continuous variables"}

# ---------------
# CORR PLOTS
# ---------------

par(mfrow = c(1,1))  # reset plotting layout

vars <- df_clean[, c("annual_kwh", "sq_meters", "dist_to_brussels", "income_euro")]
cor_matrix <- cor(vars, use = "complete.obs")
corrplot(cor_matrix,
         method = "circle",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45)

```


```{r DAGs, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Directed acyclic graphs (DAGs)", fig.width=9, fig.height=5}

library(dagitty)
library(ggdag)
library(ggplot2)
library(patchwork)

# DAG 1: Confounder
dag_confounder <- dagitty("
dag {
  building_era -> epc_label
  building_era -> annual_kwh
  epc_label -> annual_kwh
}
")

p1 <- ggdag(dag_confounder, text = TRUE, use_labels = "name") +
  theme_dag() +
  ggtitle("Building Era as a Confounder")

# DAG 2: Mediator
dag_mediator <- dagitty("
dag {
  income -> sq_meters
  sq_meters -> annual_kwh
}
")

p2 <- ggdag(dag_mediator, text = TRUE, use_labels = "name") +
  theme_dag() +
  ggtitle("Income as a Mediator")

# Combine plots
p1 + p2
```



\clearpage

## Regression Model

The initial regression model was the following:

$$
\begin{aligned}
\text{annual\_kwh}_i =\;&
\beta_0
+ \beta_1 \,\text{sq\_meters}_i
+ \beta_2 \,\text{income\_euro}_i
+ \beta_3 \,\text{occupancy\_count}_i
+ \beta_4 \,\text{dist\_to\_brussels}_i \\
&+ \boldsymbol{\beta}_5 \,\text{epc\_label}_i
+ \boldsymbol{\beta}_6 \,\text{building\_era}_i
+ \boldsymbol{\beta}_7 \,\text{roof\_color}_i
+ \beta_8 \,\text{pet\_ownership}_i \\
&+ \beta_9 (\text{income\_euro}_i \times \text{occupancy\_count}_i) \\
&+ \boldsymbol{\beta}_{10} (\text{epc\_label}_i \times \text{sq\_meters}_i) \\
&+ \boldsymbol{\beta}_{11} (\text{epc\_label}_i \times \text{occupancy\_count}_i) \\
&+ \boldsymbol{\beta}_{12} (\text{epc\_label}_i \times \text{income\_euro}_i) \\
&+ \boldsymbol{\beta}_{13} (\text{occupancy\_count}_i^2)
+ \varepsilon_i
\end{aligned}
$$

The final model retained the following predictors:

$$
\begin{aligned}
\log\text{annual\_kwh}_i =\;&
\beta_0
+ \beta_1 \,\text{sq\_meters}_i
+ \beta_2 \,\text{occupancy\_count}_i
+ \beta_3 \,\text{epc\_label}_i
+ \beta_4 \,\text{building\_era}_i
+ \varepsilon_i
\end{aligned}
$$


### Regression Results Summary Tables: Initial vs Final model


```{r echo=FALSE, warning=F, message=F}

coef_table <- summary(full_model.log)$coefficients
colnames(coef_table) <- c("Estimate", "Std. Error", "t value", "p-value")

kable(coef_table,
      digits = 5,
      caption = "Regression results for the full model (starting model)")
```


```{r echo=FALSE, warning=F, message=F}

coef_table <- summary(final_model.log_buil.era)$coefficients
ci <- confint(final_model.log_buil.era)

# Combine results
coef_table <- cbind(coef_table, ci)

# Rename columns
colnames(coef_table) <- c("Estimate", "Std. Error", "t value", "p-value", "CI 2.5%", "CI 97.5%")

kable(coef_table,
      digits = 5,
      caption = "Regression results for the final model (with 95% CI)")
```

```{r echo=FALSE, warning=F, message=F, include=FALSE}

models <- list(
  initial_model = full_model.log,
  final_model = final_model.log_buil.era
)

data.frame(
  Model = names(models),
  AIC = sapply(models, AIC),
  BIC = sapply(models, BIC),
  R2 = sapply(models, function(m) summary(m)$r.squared)
)

model_comp <- data.frame(
  Model = names(models),
  AIC = sapply(models, AIC),
  BIC = sapply(models, BIC),
  R2 = sapply(models, function(m) summary(m)$r.squared)
)
```

```{r echo=FALSE, warning=F, message=F}
kable(model_comp, digits = 3, caption = "Model comparison (initial vs. final) using AIC, BIC, and R-squared")
```



\newpage

### Diagnostic plots: Model regressing row annual energy consumption 

```{r diag.plots for model row.ann.cons, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Diagnostic plots for the selected regression model using annual energy consuption without log transformation.", fig.width=9, fig.height=8, fig.show="hold"}

par(mfrow = c(2,2))
plot(final_model)
```

\newpage

### Diagnostic plots: Model regressing log-transformed annual energy consumption 

```{r diag.plots, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Diagnostic plots for the selected regression model using log-transformed annual energy consuption.", fig.width=9, fig.height=8, fig.show="hold"}

par(mfrow = c(2,2))
plot(final_model.log_buil.era)
par(mfrow = c(1,1))

```


\newpage

# References
