---
title: "Statistical Report"
subtitle: "Determinants of Belgian Household Energy Consumption"

author: |
  GROUP 6:  
  Morodoluwa Akin-Fajiye  
  Samer Mheissen  
  Blagovesta Hristoskova  
  Jeffrey Nunoo  
  Edoardo Pennesi  
  Muthoni Wachira

date: "`r Sys.Date()`"

output:
  pdf_document:
    toc: true
    number_sections: true
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{float}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

# Libraries
library(gridExtra)
library(dplyr)

```

\newpage

# Background

Brief intro and maybe literature review

# Dataset and Variables
## Data Cleaning
```{r}
data.students <- read.table("data students.txt",header=TRUE,sep=";")
df_clean <- data.students[-c(66, 36, 135), ]
df_clean <- df_clean %>%
     filter(!is.na(epc_label) & !is.na(annual_kwh))
head(df_clean[order(-df_clean$annual_kwh), ], 10)
df_clean <- df_clean %>%
    filter(annual_kwh <= 18000)
```
# Descriptive Statistics
```{r}
library(dplyr)

numeric_vars <- df_clean %>%
  select(annual_kwh, sq_meters, income_euro, occupancy_count, dist_to_brussels)

numeric_summary <- numeric_vars %>%
  summarise(
    across(
      everything(),
      list(
        mean = ~mean(. , na.rm = TRUE),
        sd   = ~sd(. , na.rm = TRUE),
        min  = ~min(. , na.rm = TRUE),
        q1   = ~quantile(., 0.25, na.rm = TRUE),
        median = ~median(. , na.rm = TRUE),
        q3   = ~quantile(., 0.75, na.rm = TRUE),
        max  = ~max(. , na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  )

numeric_summary
```

## Categorical variables

### Summary tables

### Summary plots

- box plots

## Continuous variables

### Summary tables

### Summary plots

- box plots,
- correlation plots
- scatter plots, 
- QQ plots,

# Inferential Statistics

## Regression model

## DAGs

## Results table

| Term | Estimate | Std. Error | t value | Pr(>|t|) | adjusted p |
|------|----------|------------|---------|----------| ---------  |

- VIF analysis
- Cross validation analysis


## Diagnostic plots

- Residuals vs Fitted
- Normal Q–Q Plot
- Scale–Location Plot (Spread–Location)
- Residuals vs Leverage
- Influence (Cook’s Distance)

## Contrast analysis

- for research question 2 and 4

# Conclusions

## Research question 1

## Research question 2

## Research question 3

## Research question 4
