---
title: "Statistical Report"
subtitle: "Determinants of Belgian Household Energy Consumption"

author: |
  GROUP 6:  
  Morodoluwa Akin-Fajiye  
  Samer Mheissen  
  Blagovesta Hristoskova  
  Jeffrey Nunoo  
  Edoardo Pennesi  
  Muthoni Wachira

date: "`r Sys.Date()`"

output:
  pdf_document:
    toc: true
    number_sections: true
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{pdflscape}
  - \usepackage{float}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

# Libraries
library(gridExtra)
library(dplyr)
library(knitr)
library(kableExtra)
library(tidyr)

```

\newpage

# Background

Brief intro and maybe literature review

# Dataset and Variables
## Data Cleaning
```{r}
data.students <- read.table("data students.txt",header=TRUE,sep=";")
df_clean <- data.students[-c(66, 36, 135), ]
df_clean <- df_clean %>%
     filter(!is.na(epc_label) & !is.na(annual_kwh))
head(df_clean[order(-df_clean$annual_kwh), ], 10)
df_clean <- df_clean %>%
    filter(annual_kwh <= 18000)
```
# Descriptive Statistics
## Categorical variables
### Summary tables

These summary tables should help show if the EPC labels are balanced or skewed, if the building eras arw clustered within one period, if the roof colors or pet ownership categories have enough variation to be meaningful predictors. We can also sue them to investigate whether any category has very low counts, which can affect model stability.


```{r}
# Create summary tables for categorical variables
epc_summary <- df_clean %>%
  count(epc_label) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

building_era_summary <- df_clean %>%
  count(building_era) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

roof_color_summary <- df_clean %>%
  count(roof_color) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

pet_summary <- df_clean %>%
  count(pet_ownership) %>%
  mutate(percent = round(n / sum(n) * 100, 1))


kable(epc_summary, caption = "EPC Label Distribution") %>% kable_styling()
kable(building_era_summary, caption = "Building Era Distribution") %>% kable_styling()
kable(roof_color_summary, caption = "Roof Color Distribution") %>% kable_styling()
kable(pet_summary, caption = "Pet Ownership Distribution") %>% kable_styling()

crosscat <- df_clean %>%
  count(epc_label, building_era, roof_color, pet_ownership) %>%
  arrange(desc(n))

crosscat

```
Overall, the categorical variables show meaningful variation across categories. EPC labels are generally balanced across the different categories, indicating that extremely efficient or inefficient homes are not rare in this data. Building era and roof color are also evenly distributed while pet ownership is generally more common. None of the categories appear too sparse, suggesting that all variables can be retained for modelling without concerns about low cell counts.

### Summary plots

- box plots

## Continuous variables

### Summary tables

### Summary plots

- box plots,
- correlation plots
- scatter plots, 
- QQ plots,

# Inferential Statistics

## Regression model

## DAGs

## Results table

| Term | Estimate | Std. Error | t value | Pr(>|t|) | adjusted p |
|------|----------|------------|---------|----------| ---------  |

- VIF analysis
- Cross validation analysis


## Diagnostic plots

- Residuals vs Fitted
- Normal Q–Q Plot
- Scale–Location Plot (Spread–Location)
- Residuals vs Leverage
- Influence (Cook’s Distance)

## Contrast analysis

- for research question 2 and 4

# Conclusions

## Research question 1

## Research question 2

## Research question 3

## Research question 4
